---
title: "An Examination of Population Density's Relationship with Life Expectancy"
author: "by Ye Master of R, Zander"
output: github_document
github_document:
    toc: true
---

```{r setup, include = FALSE}
library(tidyverse)
library(geonames)
library(gapminder)
library(countrycode)
library(RColorBrewer)
knitr::opts_chunk$set(echo = TRUE)
```

We start by obtaining the information about countries' areas in square kilometers from GeoNames.orgs using a function from the geonames package. I slow the rate of iteration so as to prevent any issues with overloading their server. We then save the gapminder data from the gapminder package as an explicitly defined object in R. This latter data frame contains information about the life expectancy and population of many countries over time during recent history. 

```{r get country information, cache = TRUE}
# slow rate of scraping function
country_info_slow <- slowly(f = GNcountryInfo, 
                  rate = rate_delay(1))

# get data from GeoNames.org
country_info <- country_info_slow() %>% 
    select(countryCode, areaInSqKm) # selecting variables of interest

# load gapminder data
gapminder_data <- gapminder
```

In order to find the population density of each country, we `left_join` the country_info data to the gapminder object. We first adjust the naming convention of the gapminder data to reflect the iso convention used in the "countryCode" column of gapminder. This creates a key which the `left_join` function can use to relate the two data sets; I store the related data set as a new object. 

```{r merge data sets}
# adjust country naming convention of countries in gapminder
country_codes <- gapminder_data$country %>% 
  countrycode(origin = "country.name", destination = "iso2c")

gapminder_data <- gapminder_data %>% 
  mutate(countryCode = country_codes)

# join gapminder_data and country_info
merged_data <- gapminder_data %>% 
  left_join(country_info, by = "countryCode") %>% 
  rename(life_exp = lifeExp, population = pop, 
         area = areaInSqKm) %>% 
    # converting names to snake case
  select(country, continent, year, 
         life_exp, population, area) 
    # selecting variables of interest
```

With the two data sets now merged, we `mutate` a new "population density" variable as a function of each observation's population divided by its area. 

Calculate the population density for each observation.

```{r calculate population density}
# mutate a "pop_density" variable
merged_data <- merged_data %>% 
  mutate(area = as.numeric(area), 
    # convert area to numeric
    pop_density = population / area)
```

I graph the results twice, first as a basic scatter plot with smoothing line of population density vs. life expectancy and then another more complicated scatter broken down by continent and eached fitted with trend lines. I take a logarithmic transformation of the x-axis in both plots for enhanced visual interpretability.   

```{r basic line graph, message = FALSE}
# visualize population density versus life expectancy as a line graph
merged_data %>% 
  ggplot(aes(x = pop_density, y = life_exp), 
        position = "jitter") + # jitter to address overplotting 
    geom_point(aes(alpha = .8), show.legend = FALSE) + 
  geom_smooth(color = "white", position = "jitter") + 
    geom_point(aes(alpha = .8), show.legend = FALSE) + 
  geom_smooth(color = "white") +
  scale_x_continuous(trans = 'log10') + # logarithmic transformation
  labs(
    title = "Life Expectacy vs. Population Density",
    x = "Poplulation Density (People per km2)", 
    y = "Life Expectancy (years)") + 
  theme_dark()

```

As one can see from the graph above, it looks like population density and life expectancy are positively correlated. It appears that observations which have a greater population density also tend to have a greater life expectancy. The fitted regression line confirms this observation because of its positive slope, which increases slightly and then more sharply between a population density of 10 and 100. 

There is, however, substantial variation, which indicates that other factors may explain the life expectancy. To investigate, I visually break down the same relationship by continent below:

```{r complicated line graph, message = FALSE}
merged_data %>% 
  ggplot(aes(x = pop_density, y = life_exp),
         position = "jitter") + # jitter to address overplotting 
  geom_point(aes(color = continent, 
      alpha = .8), # make data points transparent as well
    show.legend = FALSE) + 
  geom_smooth(color = "white") +
  facet_wrap(~ continent) + 
  scale_x_continuous(trans = 'log10') + # logarithmic transformation
  labs(
    title = "Life Expectancy vs. Population Density by Continent",
    x = "Poplulation Density (People per km2)", 
    y = "Life Expectancy (years)", 
    caption = "Look at those pretty colors"
  ) + 
  scale_color_brewer(palette = "Set3") + 
  theme_dark()

```

The hypothesized general increase in life expectancy between 10 and 100 people per square kilometer holds, but there some interesting variation between continents. Life expectancy goes down somewhat noticeably in the Americas, Europe, and Oceania after around 10 people per square kilometer and then recovers soon after. Africa and Asia do not share this trend. Perhaps it marks the devastation of both World Wars, which devastated the three former continents more sharply than the latter two. This is a hypothesis that merits further investigation. 
